### 2015-07-13 narrative/data-reduction.tex
minor edits for formatting consistency.

The classical workflow where the entire dataset is written to storage for
later analysis will no longer be viable in exascale, simply because the amount of generated data will 
be too large due to capacity and performance reasons.
In the future, it will be vital to take advantage of \textit{a priori}
information when 1) writing and reading from the SSIO layer in order to gain
higher performance and predictability, and to prioritize the data that
is most useful for end users so that I/O can be finished in the time available, and 2)
performing {\em in situ} operations and analysis before storing the information.
In the fusion use case described earlier which shares commonalities with many other
DOE applications, \textit{a priori} information can indeed be provided by application
scientists regarding which data should be sent over to the SSIO layer, which we refer
to in our proposal as {\bf SIRIUS}, so that
minimally, the most science relevant data can be available for subsequent analysis.
This allows the science to be done even when the storage is in busy state - we believe
in a shared multi-user environment, it occurs often \cite{liu_hotstorage}, and gives users
high performance variability.
In this project, this capability of prioritizing data is accomplished by data refactoring, 
which includes re-organization and reductions of the data.
There are many methodologies to refactor data and the best choice will generally
be application dependent. However, our observation is that, once the choice is settled for an application, 
it will not typically change from run to run.

@@

To implement these plug-in capabilities, we will build upon our existing data staging 
capability and further allow data to be
efficiently staged, re-organized and mapped to a multi-tiered storage. Various
processing scenarios such as in-situ, in-transit or a combination of the two
will be researched and developed to meet our needs.
%The open questions that will be addressed include: 
%1) do we move the data over to
%other nodes to then reduce and re-organize, 2) do we first reduce the data and
%then later re-organize this data on different nodes?, or 3) do we use different
%cores on a node to do data-refactoring.  
Similarly we will examine the cost of reading the data. In the case where only high importance being retrieved,
overall time to move the data from SIRIUS to the
application is expected to fast. A key challenge is in the assessment of retrieving
both high and low important data that reside across storage tiers and possibly migrate to
new locations as a result of capacity pressure. There is a
clear penalty of having to discovery data (\S\ref{sec:naming-discovery}), take data in a hierarchical layout and then re-compose
this back to a single-layer view.  
%Additional research questions which must be
%addressed: Will some types of data (e.g. PIC data) be more
%conducive  for data-refactoring than others? How much semantic-domain knowledge
%will we need to get a dramatic improvement of the storage system over not doing
%anything special?

>>>>

The classical workflow where the entire dataset is written to storage for
later analysis will no longer be viable in exascale, simply because the amount of generated data will 
be too large due to capacity and performance reasons.
In the future, it will be vital to take advantage of \textit{a priori}
information when (1) writing and reading from the SSIO layer in order to gain
higher performance and predictability, and to prioritize the data that
is most useful for end users so that I/O can be finished in the time available, and (2)
performing {\em in situ} operations and analysis before storing the information.
In the fusion use case described earlier which shares commonalities with many other
DOE applications, \textit{a priori} information can indeed be provided by application
scientists regarding which data should be sent over to the SSIO layer, which we refer
to in our proposal as {\bf SIRIUS}, so that
minimally, the most science relevant data can be available for subsequent analysis.
This allows the science to be done even when the storage is in busy state - we believe
in a shared multi-user environment, it occurs often \cite{liu_hotstorage}, and gives users
high performance variability.
In this project, this capability of prioritizing data is accomplished by data refactoring, 
which includes re-organization and reductions of the data.
There are many methodologies to refactor data and the best choice will generally
be application dependent. However, our observation is that, once the choice is settled for an application, 
it will not typically change from run to run.

@@

To implement these plug-in capabilities, we will build upon our existing data staging 
capability and further allow data to be
efficiently staged, re-organized and mapped to a multi-tiered storage. Various
processing scenarios such as in-situ, in-transit or a combination of the two
will be researched and developed to meet our needs.
%The open questions that will be addressed include: 
%(1) do we move the data over to
%other nodes to then reduce and re-organize, (2) do we first reduce the data and
%then later re-organize this data on different nodes?, or (3) do we use different
%cores on a node to do data-refactoring.  
Similarly we will examine the cost of reading the data. In the case where only high importance being retrieved,
overall time to move the data from SIRIUS to the
application is expected to fast. A key challenge is in the assessment of retrieving
both high and low important data that reside across storage tiers and possibly migrate to
new locations as a result of capacity pressure. There is a
clear penalty of having to discovery data (\S\ref{sec:naming-discovery}), take data in a hierarchical layout and then re-compose
this back to a single-layer view.  
%Additional research questions which must be
%addressed: Will some types of data (e.g. PIC data) be more
%conducive  for data-refactoring than others? How much semantic-domain knowledge
%will we need to get a dramatic improvement of the storage system over not doing
%anything special?

### 2015-07-13 narrative/data-reorg.tex
minor edits for formatting consistency.

\paragraph{Proposed research approach:}
General methods to re-organize the data fall into the following categories
\begin{itemize}
\item Precision based re-organization. This is where the most significant bytes of the data are
all group together from each object. This data generally will have a higher utility then the data with
the least significant bytes.  This process has the following three steps which need to be
efficiently implemented when data gets placed to the storage system and from the storage 
system. 
%
First, the data needs to be re-arranged which involves memory operations and needs
to be done in-situ. This operation involves no communication.  Data needs to be copied, so
the memory requirements are increased. At most this requires two copies an individual dataset
in memory, and we will investigate techniques to allocate and deallocate this memory if the user
will specify that the data will be overwritten after it is written to the storage system. This is often
the case for many of the data quantities written form the simulation, but there are many cases
where, for example, we want to write all of the particles from a Particle In Cell (PIC) simulation. 
Since the particles will be used later in the calculation, we need to duplicate the storage. Our
observation with working with the XGC1, GTC, Warp, and PiconGPU simulations (all leadership
class simulations) is that we can temporary increase the storage of the particles, and then 
release them since the temporary arrays used in the calculations are often freed when a PIC
iteration is finished.. 
%
The second step to give the metadata that allows the system to understand that an object is now contained in multiple bins of data. One of the challenges is the ability to bring together multiple bins
of data together (the high priority data along with the medium priority data) efficiently when the 
data is being read.  One research question is how we can manage the metadata efficiently.
We propose to place this knowledge into the middleware which will communicate this to the 
storage system.  This will give the storage system potential ways to increase the concurrency 
when accessing this information. For example, we can place the highest priority bits of an object
on the fastest storage system, and the next level of importance in the campaign storage. If a 
user request precision to a level where we need to access both, concurrently we can be
accessing this data, and moving the data to memory where it can be re-arranged. 
%
Since data will be placed in different bins along the hierarchy we will need to understand how
to initially place this in the fastest storage, and then evict this from the fastest storage without
effecting the performance of the simulation which is generating this data. Policies must be 
placed to ensure that we do not create any internal interference during these operations. We
will investigate doing this with ``meta-bots'' first introduced in the LWFS project and have 
these bots move the data.{\bf probably belongs in a different section}

@@

These operations also require data transformations (for example an FFT and an inverse FFT)
and then re-arranging the data in real-space once we re-organize this. The research questions
we must address are: 1) what is the cost of re-organzing the data, 2) how do we

>>>>

\paragraph{Proposed research approach:}
General methods to re-organize the data fall into the following categories
\begin{itemize}
\item Precision based re-organization. This is where the most significant bytes of the data are
all group together from each object. This data generally will have a higher utility then the data with
the least significant bytes.  This process has the following three steps which need to be
efficiently implemented when data gets placed to the storage system and from the storage 
system. 
%
First, the data needs to be re-arranged which involves memory operations and needs
to be done in situ. This operation involves no communication.  Data needs to be copied, so
the memory requirements are increased. At most this requires two copies an individual dataset
in memory, and we will investigate techniques to allocate and deallocate this memory if the user
will specify that the data will be overwritten after it is written to the storage system. This is often
the case for many of the data quantities written form the simulation, but there are many cases
where, for example, we want to write all of the particles from a Particle In Cell (PIC) simulation. 
Since the particles will be used later in the calculation, we need to duplicate the storage. Our
observation with working with the XGC1, GTC, Warp, and PiconGPU simulations (all leadership
class simulations) is that we can temporary increase the storage of the particles, and then 
release them since the temporary arrays used in the calculations are often freed when a PIC
iteration is finished.. 
%
The second step to give the metadata that allows the system to understand that an object is now contained in multiple bins of data. One of the challenges is the ability to bring together multiple bins
of data together (the high priority data along with the medium priority data) efficiently when the 
data is being read.  One research question is how we can manage the metadata efficiently.
We propose to place this knowledge into the middleware which will communicate this to the 
storage system.  This will give the storage system potential ways to increase the concurrency 
when accessing this information. For example, we can place the highest priority bits of an object
on the fastest storage system, and the next level of importance in the campaign storage. If a 
user request precision to a level where we need to access both, concurrently we can be
accessing this data, and moving the data to memory where it can be re-arranged. 
%
Since data will be placed in different bins along the hierarchy we will need to understand how
to initially place this in the fastest storage, and then evict this from the fastest storage without
effecting the performance of the simulation which is generating this data. Policies must be 
placed to ensure that we do not create any internal interference during these operations. We
will investigate doing this with ``meta-bots'' first introduced in the LWFS project and have 
these bots move the data.{\bf probably belongs in a different section}

@@

These operations also require data transformations (for example an FFT and an inverse FFT)
and then re-arranging the data in real-space once we re-organize this. The research questions
we must address are: (1) what is the cost of re-organzing the data, (2) how do we

### 2015-07-13 narrative/intro.tex
minor edits for formatting consistency.

\begin{table}[htbp]
\small
\vspace{2ex}
\begin{center}
\caption {Relevance to SSIO}
\label{tab:call}
\begin{tabular}{ | p {0.2cm} | p{5cm} | p{7.5cm} | p{2.2cm} |}\hline
\multicolumn{2}{|c|}{\bf Call area} & \multicolumn{1}{|c|}{\bf Relevant SIRIUS work} & \multicolumn{1}{|c|}{\bf Proposed research} \\\hline\hline
1 & Data abstractions & Create new APIs  and annotations to guide data placement across the storage system & \S\ref{sec:data-description} \\\hline
2 & Mapping science data models onto hierarchical storage &  Enable science-driven mapping, re-organization, reduction and prioritization of data chunks derive from application objects & \S\ref{sec:data-refactor}, \S\ref{sec:data-description},  \S\ref{sec:managing-data-life}\\\hline
3 & Mechanisms for data movement across the storage hierarchy & Enable science-driven data placement and migration based on policies& \S \ref{sec:managing-data-life},  \S\ref{sec:migration} \\\hline
4 &  Extend the interaction of middleware with the storage system & Manage data placement and migration between the application, middleware and storage system using an autonomic cross-layer approach & \S\ref{sec:managing-data-life}, \S\ref{sec:qos}, \S\ref{sec:naming-discovery}  \\\hline
5 & Design and implement of I/O middleware architectures  & Create APIs to guide the placement of chunks of data across the SSIO and autonomically move data& \S\ref{sec:data-description}, \S\ref{sec:managing-data-life}, \S\ref{sec:qos}, \S\ref{sec:naming-discovery} \\ \hline
6 &Approaches to improve the ability of SSIO software to support checkpoint/restart &  We will have utility functions defined for the lifetime of data, and allow CR data to be automatically placed in the fastest storage and then later purged & \S\ref{sec:data-description}, \S\ref{sec:managing-data-life}, \S\ref{sec:qos}, \S\ref{sec:data-description} \\ \hline
\end{tabular}
\end{center}
\vskip -0.5cm
\end{table}

@@

\paragraph{Simulation Perspective}
\label{subsec:sim-perspective}
A simulation scientist should ideally be able to
obtain an estimate of how much time will be required to write data,
and how much storage space is available during the simulations lifetime.
Further, they would like Quality of Service (QoS) assurances, and that
issues such as varying bandwidth can be handled appropriately and 
automatically. For example, a set of user-specified rules would allow 
the system to make autonomic decisions at runtime
~\S~\ref{sec:managing-data-life}~\S~\ref{sec:qos}.
%
Our approach is based on giving users the ability to define the utility of
their data \S\ref{sec:data-description}. This includes the ability to assign
varying levels of importance to individual data objects in their entirety,
or to specific subsets and pieces. Key research challenges include: 1)
developing an expressive and easy to use a program interface, 2) formulating
a system model that can support autonomic decision, for example, for saving
data based on user-defined importance and utility of
data~\S~\ref{sec:managing-data-life}), and 3) defining the interface between
the system and the back-end storage for both single and multiple running
applications~\S~\ref{sec:qos}.

@@

\paragraph{Storage System Perspective}
\label{subsec:storage-perspective}
From the perspective of the storage system, it is necessary to manage a
single I/O request from a user in the context of all requests from all
users, and try to optimize the entire system accordingly. This entails
providing quality of service (~\S\ref{sec:qos} to the user to minimize
resource contention, distribute the data across all of the tiers of storage
in a known consistent manner (\S~\ref{sec:init-plac-data}), and automatically move data from one part of
the storage hierarchy to another (~\ref{sec:migration}).

One of our research objectives is to understand how the
storage system can interact with the middleware   so
that the data with the highest utility is kept on the fastest layers of the
storage system for as long as specified. We need to maintain 
QoS and therefore, be able to provide fuzzy estimates of the time it will take to write/read data from
the different layers to inform any migration/eviction decisions, and it
is also critical that the storage system can manage (discover and retrieve) the data across all of the
layers automatically, without user intervention. 
Our approach here is that the storage system needs to define bounds of time estimates
(min time, max time), and as our system evolves, we want to allow plug-ins which can help
reduce the range of the bounds.
Overall, our goal is to enable predictable performance acorss the storage system, and we will develop mechanisms to enable this
~\S\ref{sec:naming-discovery}.

@@

From the reading perspective, the data must be accessible in a fast and
predictable manner for all read access patterns, with different access
patterns having their own requirements for latency, throughput and
consistency. Here the user must be able to describe the read patterns and
query for performance estimations (~\S~\ref{sec:data-description}), be able
to quickly access the most important data (\S~\ref{sec:data-refactor}) and recieve
performance gurantees (~\S\ref{sec:qos}) on the time required to complete
the read request. The challenge here is to address the variety of read
patterns that drive knowledge discover.

>>>>

\begin{table}[htbp]
\small
\vspace{2ex}
\begin{center}
\caption {Relevance to SSIO}
\label{tab:call}
\begin{tabular}{ | p {0.2cm} | p{5cm} | p{7.5cm} | p{2.2cm} |}\hline
\multicolumn{2}{|c|}{\bf Call area} & \multicolumn{1}{|c|}{\bf Relevant SIRIUS work} & \multicolumn{1}{|c|}{\bf Proposed research} \\\hline\hline
1 & Data abstractions & Create new APIs  and annotations to guide data placement across the storage system & \S\ref{sec:data-description} \\\hline
2 & Mapping science data models onto hierarchical storage &  Enable science-driven mapping, re-organization, reduction and prioritization of data chunks derive from application objects & \S\ref{sec:data-refactor}, \S\ref{sec:data-description}, \S\ref{sec:managing-data-life}\\\hline
3 & Mechanisms for data movement across the storage hierarchy & Enable science-driven data placement and migration based on policies& \S\ref{sec:managing-data-life}, \S\ref{sec:migration} \\\hline
4 &  Extend the interaction of middleware with the storage system & Manage data placement and migration between the application, middleware and storage system using an autonomic cross-layer approach & \S\ref{sec:managing-data-life}, \S\ref{sec:qos}, \S\ref{sec:naming-discovery}  \\\hline
5 & Design and implement of I/O middleware architectures  & Create APIs to guide the placement of chunks of data across the SSIO and autonomically move data& \S\ref{sec:data-description}, \S\ref{sec:managing-data-life}, \S\ref{sec:qos}, \S\ref{sec:naming-discovery} \\ \hline
6 &Approaches to improve the ability of SSIO software to support checkpoint/restart &  We will have utility functions defined for the lifetime of data, and allow CR data to be automatically placed in the fastest storage and then later purged & \S\ref{sec:data-description}, \S\ref{sec:managing-data-life}, \S\ref{sec:qos}, \S\ref{sec:data-description} \\ \hline
\end{tabular}
\end{center}
\vskip -0.5cm
\end{table}

@@

\paragraph{Simulation Perspective}
\label{subsec:sim-perspective}
A simulation scientist should ideally be able to
obtain an estimate of how much time will be required to write data,
and how much storage space is available during the simulations lifetime.
Further, they would like Quality of Service (QoS) assurances, and that
issues such as varying bandwidth can be handled appropriately and 
automatically. For example, a set of user-specified rules would allow 
the system to make autonomic decisions at runtime
\S\ref{sec:managing-data-life} \S\ref{sec:qos}.
%
Our approach is based on giving users the ability to define the utility of
their data \S\ref{sec:data-description}. This includes the ability to assign
varying levels of importance to individual data objects in their entirety,
or to specific subsets and pieces. Key research challenges include: (1)
developing an expressive and easy to use a program interface, (2) formulating
a system model that can support autonomic decision, for example, for saving
data based on user-defined importance and utility of
data \S\ref{sec:managing-data-life}), and (3) defining the interface between
the system and the back-end storage for both single and multiple running
applications \S\ref{sec:qos}.

@@

\paragraph{Storage System Perspective}
\label{subsec:storage-perspective}
From the perspective of the storage system, it is necessary to manage a
single I/O request from a user in the context of all requests from all
users, and try to optimize the entire system accordingly. This entails
providing quality of service (\S\ref{sec:qos} to the user to minimize
resource contention, distribute the data across all of the tiers of storage
in a known consistent manner (\S\ref{sec:init-plac-data}), and automatically move data from one part of
the storage hierarchy to another (\S\ref{sec:migration}).

One of our research objectives is to understand how the
storage system can interact with the middleware   so
that the data with the highest utility is kept on the fastest layers of the
storage system for as long as specified. We need to maintain 
QoS and therefore, be able to provide fuzzy estimates of the time it will take to write/read data from
the different layers to inform any migration/eviction decisions, and it
is also critical that the storage system can manage (discover and retrieve) the data across all of the
layers automatically, without user intervention. 
Our approach here is that the storage system needs to define bounds of time estimates
(min time, max time), and as our system evolves, we want to allow plug-ins which can help
reduce the range of the bounds.
Overall, our goal is to enable predictable performance acorss the storage system, and we will develop mechanisms to enable this
\S\ref{sec:naming-discovery}.

@@

From the reading perspective, the data must be accessible in a fast and
predictable manner for all read access patterns, with different access
patterns having their own requirements for latency, throughput and
consistency. Here the user must be able to describe the read patterns and
query for performance estimations (\S\ref{sec:data-description}), be able
to quickly access the most important data (\S\ref{sec:data-refactor}) and recieve
performance gurantees (\S\ref{sec:qos}) on the time required to complete
the read request. The challenge here is to address the variety of read
patterns that drive knowledge discover.
